---
title: "Beta distribution"
author: "Bruno Melati"
date: "2025-08-20"
output: pdf_document
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To set the costs and benefits parameters we chosed a **beta distribution**. This distribution was chosed for some reasons:

  - The values are all positive. As do not exist negative cost or negative benefit, it fit to our data;
  - The values are defined on the inverval [0,1]. It could be used as the percentage of costs and benefits;
  - As the values are bounded between 0 and 1, we know the maximum value of benefits for one species, that is $B \le k_{i}$, where *k* is the degree of *i*;
  - The shape of the curve assume diffent forms as a function of the paramters. In nature we really do not know the shape of the distribuitions of costs and benefits;
  - It have a solution for ou case, which the other option of distribution (lognormal) do not.

The beta distribution have two parameters $\alpha$ abd $\beta$ and both control the shpae of the distribution. Therefore, 

$$ X \approx Beta(\alpha,\beta)$$

where, $\alpha > 0$ and  $\beta > 0$. Both parameters control de "influence" of of the shape bias. Higher $\alpha$ values, higher the concentration towards the right side of the curve. Higher $\beta$ values, higher the concentration towards the left side of the curve. 

The mean and variance of Beta depends on $\alpha$ and $\beta$ parameters, as:

$$\bar{x} = \frac{\alpha}{\alpha + \beta}$$
$$\mathrm{Var}(X) = \frac{\alpha \beta} {(\alpha + \beta)^2 (\alpha + \beta + 1)}$$

and $$E[X] = \frac{\alpha}{\alpha + \beta}$$

```{r}
 # Define the function to integrate
    func_to_integrate <- function(x, alpha, beta) {
      x * dbeta(x, shape1 = alpha, shape2 = beta)
    }

    # Set your alpha and beta values
    alpha_val <- 2
    beta_val <- 3

    # Perform the numerical integration
    integration_result <- integrate(func_to_integrate, lower = 0, upper = 1, alpha = alpha_val, beta = beta_val)

    # The expected value is the 'value' element of the result
    expected_value <- integration_result$value
    print(expected_value)

```




















